---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gmRi

<!-- badges: start -->
<!-- badges: end -->

The goal of gmRi is to consolidate useful reasearch tools under one consistent repository. Things like official GMRI colors or stylesheets for Rmarkdown documents are easy first steps in creating consistent and professsional looking documents. Consistent workflows for creating datasets from outside sources like from satellite SST or NERACOOS buoy arrays will make our work take less time and be consistent in-house.

## Installation

You can install the development version of gmRi from [GitHub](www.github.com) with:

``` r
devtools::install_.packages_github("https://github.com/gulfofmaine/gmRi")
```

## Example

This is a basic example of how to pull GMRI colors for a ggplot2 figure:

```{r example}
library(gmRi)
library(ggplot2)
## basic example code

ggplot(mtcars, aes(hp, mpg)) +
  geom_point(color = gmri_cols("gmri blue"), size = 4, alpha = .8)
```

If you want to use a gmri color palette you can access them this way:

```{r}
ggplot(mpg, aes(manufacturer, fill = manufacturer)) +
   geom_bar() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_fill_gmri(palette = "mixed", guide = "none")
```

